The Companion Mass Distribution to the ELM WD Sample

### 2014-10-04 documents/letter/ms.tex
Updated intro, added references

% Affiliations
\newcommand{\colum}{1}
\altaffiltext{\colum}{Department of Astronomy, 
		              Columbia University, 
		              550 W 120th St., 
		              New York, NY 10027, USA}
% \altaffiltext{\adrn}{To whom correspondence should be addressed: adrn@astro.columbia.edu}

@@

%\keywords{
%  cosmology: dark matter
%  ---
%  Galaxy: structure
%  ---
%  Galaxy: halo
%}

%\begin{figure*}[!h]
%\begin{center}
%\includegraphics[width=\textwidth]{PATH}
%\caption{ CAPTION }\label{fig:FIGNAME}
%\end{center}
%\end{figure*}

@@

The dramatic increase in the number of spectroscopically confirmed white dwarfs (WDs) is seen in the most recently compiled sample from the Data Release 7 (DR7) in the Sloan Digital Sky Survey (SDSS), containing some 20,000 confirmed WDs (Kleinman et al 2013). While the distribution of WD masses is strongly peaked at $0.6~\Msun$, a small but non-negligible number (6\%) of low mass WDs was observed, with $M<0.4~\Msun$.

These Low Mass WDs (LMWDs) comprise an astrophysically unique population. Except in special cases of extreme metallicity (Kilic et al 2007), the Galaxy is not old enough to produce WDs with mass $<0.5~\Msun$ through single star evolution. In their seminal work, Marsh, Dhillon, \& Duck (1995) discovered companions around five LMWDs, cementing the requirement that these objects be formed through binary evolution. In addition to SDSS, LMWDs have been observed in similar numbers in the Palomar Green Survey (Liebert et al 2005).

Using a reduced proper motion selection criterion, the Extrememly Low Mass (ELM) WD Survey specifically searched for LMWDs in SDSS. After their initial investigation found 11 of 12 LMWDs had close binary companions (Brown et al. 2010), further radial velocity follow-up has now brought the sample of ELM WDs to 61 systems, 55 of which have measured radial velocity variations (Kilic et al 2011, 2012, Brown et al 2012, 2013).

While the orbital periods and radial velocities indicated the companions were most likely WDs, it was realized that these LMWDs could have NS companions (van Leeuwen et al 2007). However this and other radio searches for pulsed emission were fruitless (Agueros et al 2009a, 2009b). X-ray observations provide more stringent limits on the presence of a NS, and searches using {\it XMM-Newton} and {\it Chandra} have all yielded non-detections. Yet, such systems must exist since LMWDs are observed as companions around millisecond pulsars. Because of the resources required for these targeted observations, selecting the most likely candidates for multiwavelength follow-up is crucial. We outline the standard technique below.

@@

\begin{equation}
	m_f = \frac{M_2^3}{\left(M_1+M_2\right)^2} {\rm sin}^3 i = \frac{P_{\rm orb}}{2\pi G} K^3 \label{eq:massfunc}
\end{equation}

The righthand side of this equation is the well-known mass function. $M_2$ is minimized for an edge-on orbit, with $i = 90^{\circ}$. For any individual system, deriving the likelihood of a $M_2$ for that system ($P(M_2)$) involves solving Equation 1 for $M_2$ for different values for $i$, weighting the inclination angle by sin $i$. A basic calculation for the probability that the companion to any individual system is a NS can therefore be calculated:

@@

Where $i_{\rm NS}$ is the inclination angle from Equation 1 when $M_2$ is the mass of a NS, typically $\approx1.35~\Msun$. However, implicit in this calculation is that all companion masses are equally likely a priori. Clearly this is not the case as evidenced by e.g. SDSS (kleinman et al 2013) who show the WD mass distribution is strongly peaked around $0.6~\Msun$. Including this as a prior on the companion mass does not improve the situation since it is unlikely that the mass distribution of single WDs and WDs in binaries is the same. Equation 2 should therefore be used only as an approximation for the NS probability of a companion.

With 55 systems with precise $M_1$, $P_{\rm orb}$, and $K$, the population is large enough to be modeled. Furthermore, since the population was selected based on characteristics determined by the primary mass alone, at least with regard to the inclination angle and companion mass, the population is unbiased. In this work we develop a model to determine the mass distribution of companions to ELM WDs in an attempt to answer several specific scientific questions: 1.\ Can the population as a whole be modeled using a simple description of the companion masses (i.e. one Gaussian)? 2.\ How many NSs are implied by the distribution of $M_{2,min}$'s? 3.\ Can we determine the likelihood of any individual system including a NS companion, based on the most likely models? 4.\ Are the six systems without detected radial velocity variations consistent with the population as a whole, or are these a fundamentally separate population (i.e. single ELM WDs)? 7.\ Can our model place limits on the space density of ELM WD binaries with NS companions?

@@

%\begin{figure}[h!]
%\begin{center}
%\includegraphics[width=0.7\columnwidth]{figures/Porb_M1/Porb_M1.jpg}
%\caption{The $P_{\rm orb}$ - $M_1$ distribution of the ELM WD sample. The three eclipsing systems, with known $M_2$ are shown by the filled circles. Arrows indicate the masses of ELM WDs without detected radial velocity variation. Triangles indicate the positions of the known WD-NS binaries. From $P_{\rm orb}$ and $M_1$ alone, the two populations are indistinguishable.}
%\end{center}
%\end{figure}

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[width=1\columnwidth]{model_post.eps}
\caption{Replace this text with your caption}
\end{center}
\end{figure}

@@

In the ELM sample three systems are eclipsing binaries, with known companion masses (J0345$+$1748: $M_2=0.72~\Msun$, J0651$+$2844: $M_2=0.50~\Msun$, and J0751$-$0141: $M_2=0.97~\Msun$). These are taken into account in our model through equation {\bf XXX}. Since the inclination angle cannot vary, these systems will strongly affect the likelihood of any particular model.

@@

We apply our model to the ELM WD data set, using 10 indepedent Markov chains, each run for 100,000 steps. The acceptance fraction is {\bf XXX}. We determine an autocorrelation length of {\bf XXX} steps, and correspondingly, only include every {\bf XXX} samples for our posterior distributions. {\bf Is there anything else we need to say about the applying the method to the data set?}

@@

How does the companion mass distribution compare to the single WD mass distribution? Based on their population of $\sim$ 14,000 spectroscopically confirmed DA WDs in SDSS, Kleinman et al (2013) found that WDs are well approximated by a four component Gaussian mixture model. Their mixture model is dominated by two Gaussians with means $\sim0.6~\Msun$, containing over 75\% of all WDs. Figure {\bf XXX} shows that the best fit Gaussian to our distribution of WD companion masses is also centered around $\sim0.6!\Msun$. However, our distribution is significantly wider ($\sigma \approx 0.3~\Msun$ compared with $\sigma \approx 0.1~\Msun$). This could be due to our simplistic assumption of a single Gaussian mass distribution. A more complicated model would include multiple Gaussians with varying weights, however adding only one more Gaussian component doubles the number of free parameters from three to six. Finding the maximum likelihood in a three dimensional parameter space is computationally expensive, and moving to a six dimensional space may require more complex numerical routines.

There is no reason to expect the true companion mass distribution to ELM WDs follows a Gaussian curve. However, the posterior distribution in $\mu$ and $\sigma$ is well behaved, with a clear high probability region, similar to the posterior distributions for $\mu$ and $\sigma$ for our first two test cases shown in Figure {\bf XXX}. The part of the distribution with a low $\mu$ and high $\sigma$ is indicative of a low mass WD component. A low mass component is expected from binary evolution theory; recent population synthesis work predicts more than half of all He WDs in DWDs will have He WD companions (Toonen et al 2012). Our results here suggest that this is not the case, our posterior suggests a Gaussian distribution centered around $0.6~\Msun$, firmly in the mass range for CO WDs. This suggests that our results on the companion masses of the secondaries in these systems could constrain population synthesis models. We plan on exploring this in a future work.

Systems with NS companions cannot be differentiated from those with WD companions through orbital period and ELM WD mass alone. Our model attempts to suggest a NS fraction for the population as a whole. This is an improvement over NS probabilities for individual companions calculated through equation {\bf XXX}. The third panel in Figure {\bf XXX} indicates a NS fraction of a few percent. For 55 systems with observed radial velocity variations, this corresponds to a 1-2 systems with NS companions. However, the top left panel in Figure {\bf XXX} shows that even a system with $M_{2,min}=1.0~\Msun$ can be a WD from the high mass tail of a Gaussian distribution. Furthermore, in their Gaussian mixture fit to the SDSS DA WD mass distribution, Kleinman et al (2013) found a high mass WD component extending above $1~\Msun$. Our single Gaussian model does not include this high mass component. It is not guaranteed that any ELM WDs have NS companions. Nevertheless, a quite firm result of our model is an upper limit on the fraction of ELM WDs with NS companions of {\bf XXXX}\%.

If we combine our result of a NS fraction of {\bf XXX} with the ELM WD space density as determined from the Palomar Green Survey: $4\times10^{-14}$pc$^{-3}$yr$^{-1}$ (Liebert et al 2005) {\bf Is there a more recent estimate?} we can get a space density of ELM WD + NS binaries. This suggests that there are {\bf XXX} such binaries within a few kpc of us. Compare to the number of known MSP+LMWDs.

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[width=1\columnwidth]{ELM_post.eps}
\caption{Replace this text with your caption}
\end{center}
\end{figure}

@@

>>>>

% Affiliations
\newcommand{\colum}{1}
\altaffiltext{\colum}{Department of Astronomy, 
		              Columbia University, 
		              550 W 120th St., 
		              New York, NY 10027, USA}

@@

@@

The dramatic increase in the number of spectroscopically confirmed white dwarfs (WDs) is seen in the most recently compiled sample from the Data Release 7 (DR7) in the Sloan Digital Sky Survey (SDSS), containing some 20,000 confirmed WDs \citet{kleinman13}. While the distribution of WD masses is strongly peaked at $0.6~\Msun$, a small but non-negligible number (6\%) of low mass WDs (LMWD) was observed, with $M<0.4~\Msun$. In addition to SDSS, LMWDs have been observed in similar numbers elsewhere \citep[e.g. the Palomar Green Survey][]{liebert05}.

Except in special cases of extreme metallicity \citep{kilic07}, the Galaxy is not old enough to produce these LMWD through single star evolution. These objects are formed through dynamical interactions with another star, in which the primary overfills its Roche lobe in either an unstable, common envelope \citep{vdSluys06}, or a phase of stable but non-conservative mass transfer \citep{woods12}. After the primary has become a WD, the secondary overfills its Roche lobe, forming a common envelope prior to core helium ignition, leaving a helium core WD secondary in a tight orbit with either a He- or CO- core WD \citep{han98}.

In their seminal work, \citet{marsh95} discovered companions around five LMWDs in their sample of seven objects, cementing LMWDs as fruitful targets for close binary searches. Using a reduced proper motion selection criterion, the Extremely Low Mass (ELM) WD Survey specifically searched for LMWDs in SDSS using a photometric constraint combined with a reduced proper motion constraint. After their initial investigation found 11 of 12 LMWDs had close binary companions \citep{ELMI}, further radial velocity follow-up has now brought the sample of ELM WDs to 61 systems, 55 of which have measured radial velocity variations \citep{ELMII, ELMIII, ELMIV, ELMV}.

While the orbital periods and radial velocities indicated the companions were most likely WDs, it was realized that these LMWDs could have NS companions \citep{vLeeuwen07}. However this and other radio searches for pulsed emission, as well as searches for blackbody X-ray emission using {\it XMM-Newton} and {\it Chandra}, have all been fruitless \citet{agueros09a,agueros09b,kilic13}. Yet, such systems must exist since LMWDs are observed as companions around millisecond pulsars. Since the orbital solutions provided by the ELM survey combined with the precision of radio data would overconstrain the system, finding even one system would be extraordinarily interesting. The NS binaries with WD companions bright enough to detect spectroscopic radial velocity variations have precisely determined NS masses, allowing for an understanding of the binary evolution processes forming these systems \citep{vKerkwijk96,callanan98,bassa06,antoniadis12}. The resources required for these targeted observations, selecting the most likely candidates for multiwavelength follow-up is crucial. We outline the standard technique below.

@@

\begin{equation}
	\frac{M_2^3}{\left(M_1+M_2\right)^2} {\rm sin}^3 i = \frac{P_{\rm orb}}{2\pi G} K^3 = m_f \label{eq:massfunc}
\end{equation}

The righthand side of this equation is the well-known mass function ($M_f$). $M_2$ is minimized for an edge-on orbit, with $i = 90^{\circ}$. For any individual system, deriving the likelihood of a $M_2$ for that system ($P(M_2)$) involves solving Equation 1 for $M_2$ for different values for $i$, weighting the inclination angle by sin $i$. A basic calculation for the probability that the companion to any individual system is a NS can therefore be calculated:

@@

Where $i_{\rm NS}$ is the inclination angle from Equation 1 when $M_2$ is the mass of a NS, typically $\approx1.35~\Msun$. However, implicit in this calculation is that all companion masses are equally likely a priori. Clearly this is not the case as evidenced by \citep{kleinman13} who show the WD mass distribution in SDSS is strongly peaked around $0.6~\Msun$. Including this as a prior on the companion mass does not improve the situation since it is unlikely that the mass distribution of single WDs and WDs in binaries is the same. Equation 2 should therefore be used only as an approximation for the NS probability of a companion.

With 55 systems with precise $M_1$, $P_{\rm orb}$, and $K$, the ELM sample is large enough to be statistically modeled. Furthermore, since the population was selected based on characteristics determined by the primary mass alone, at least with regard to the inclination angle and companion mass, the population is unbiased. In this work we develop a model to determine the mass distribution of companions to ELM WDs in an attempt to answer several specific scientific questions: 1.\ Can the population as a whole be modeled using a simple description of the companion masses? 2.\ How many NSs does our simplistic model imply? 3.\ Can we determine the improved likelihood of any individual system having a NS companion, taking into account the overall rate of NS's in the sample implied by our model? 4.\ Are the six systems without detected radial velocity variations consistent with the population as a whole, or are these a fundamentally separate population (i.e. single ELM WDs)? 5.\ Can our model place limits on the space density of ELM WD binaries with NS companions?

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[angle=90,width=0.7\columnwidth]{Porb_M1.eps}
\caption{The $P_{\rm orb}$ - $M_1$ distribution of the ELM WD sample. The three eclipsing systems, with known $M_2$ are shown by the filled circles. Arrows indicate the masses of ELM WDs without detected radial velocity variation. Triangles indicate the positions of the known WD-NS binaries. From $P_{\rm orb}$ and $M_1$ alone, the two populations are indistinguishable.}
\end{center}
\end{figure}

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[angle=90,width=1\columnwidth]{model_post.eps}
\caption{Replace this text with your caption}
\end{center}
\end{figure}

@@

In the ELM sample three systems are eclipsing binaries, with known companion masses (J0345$+$1748: $M_2=0.72~\Msun$, J0651$+$2844: $M_2=0.50~\Msun$, and J0751$-$0141: $M_2=0.97~\Msun$) {\bf Cite something here}. These are taken into account in our model through equation {\bf XXX}. Since the inclination angle cannot vary, these systems will strongly affect the likelihood of any particular model.

@@

We apply our model to the ELM WD data set, using 10 independent Markov chains, each run for 100,000 steps. The acceptance fraction is {\bf XXX}. We determine an autocorrelation length of {\bf XXX} steps, and correspondingly, only include every {\bf XXX} samples for our posterior distributions. {\bf Is there anything else we need to say about the applying the method to the data set?}

@@

How does the companion mass distribution compare to the single WD mass distribution? Based on their population of $\sim$ 14,000 spectroscopically confirmed DA WDs in SDSS, \citet{kleinman13} found that WDs are well approximated by a four component Gaussian mixture model. Their mixture model is dominated by two Gaussians with means $\sim0.6~\Msun$, containing over 75\% of all WDs. Figure {\bf XXX} shows that the best fit Gaussian to our distribution of WD companion masses is also centered around $\sim0.6!\Msun$. However, our distribution is significantly wider ($\sigma \approx 0.3~\Msun$ compared with $\sigma \approx 0.1~\Msun$). This could be due to our simplistic assumption of a single Gaussian mass distribution. A more complicated model would include multiple Gaussians with varying weights, however adding only one more Gaussian component doubles the number of free parameters from three to six. Finding the maximum likelihood in a three dimensional parameter space is computationally expensive, and moving to a six dimensional space may require more complex numerical routines.

There is no reason to expect the true companion mass distribution to ELM WDs follows a Gaussian curve. However, the posterior distribution in $\mu$ and $\sigma$ is well behaved, with a clear high probability region, similar to the posterior distributions for $\mu$ and $\sigma$ for our first two test cases shown in Figure {\bf XXX}. The part of the distribution with a low $\mu$ and high $\sigma$ is indicative of a low mass WD component. A low mass component is expected from binary evolution theory; recent population synthesis work predicts more than half of all He WDs in DWDs will have He WD companions \citet{toonen12}. Our results here suggest that this is not the case, our posterior suggests a Gaussian distribution centered around $0.6~\Msun$, firmly in the mass range for CO WDs. This suggests that our results on the companion masses of the secondaries in these systems could constrain population synthesis models. We plan on exploring this in a future work.

Systems with NS companions cannot be differentiated from those with WD companions through orbital period and ELM WD mass alone. Our model attempts to suggest a NS fraction for the population as a whole. This is an improvement over NS probabilities for individual companions calculated through equation {\bf XXX}. The third panel in Figure {\bf XXX} indicates a NS fraction of a few percent. For 55 systems with observed radial velocity variations, this corresponds to a 1-2 systems with NS companions. However, the top left panel in Figure {\bf XXX} shows that even a system with $M_{2,min}=1.0~\Msun$ can be a WD from the high mass tail of a Gaussian distribution. Furthermore, in their Gaussian mixture fit to the SDSS DA WD mass distribution, \citet{kleinman13} found a high mass WD component extending above $1~\Msun$. Our single Gaussian model does not include this high mass component. It is not guaranteed that any ELM WDs have NS companions. Nevertheless, a quite firm result of our model is an upper limit on the fraction of ELM WDs with NS companions of {\bf XXXX}\%.

If we combine our result of a NS fraction of {\bf XXX} with the ELM WD space density as determined from the Palomar Green Survey: $4\times10^{-14}$pc$^{-3}$yr$^{-1}$ \cite{liebert05} {\bf Is there a more recent estimate?} we can get a space density of ELM WD + NS binaries. This suggests that there are {\bf XXX} such binaries within a few kpc of us. Compare to the number of known MSP+LMWDs.

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[angle=90,width=1\columnwidth]{ELM_post.eps}
\caption{Replace this text with your caption}
\end{center}
\end{figure}

@@

\bibitem[Goodman~\&\ Weare(2010)]{goodman10}
Goodman,~J. \& Weare,\ J.,
2010, Comm.\ App.\ Math.\ Comp.\ Sci., 5, 65
